Improved state-space models for inference about spatial and temporal variation in abundance from count data

### 2012-09-06 Dail_Mad_ext.tex
expanding the description of the Dail-madsen model

%\subsection{Original Formulation}

As originally formulated, the DM model includes submodels for four
conditionally related processes: the initial abundance state, apparent
survival, recruitment, and detection probability. \citet{dail_madsen:2011}
proposed two distributions for initial abundance, the Poisson and negative
binomial:
\begin{gather}
N_{it} \sim \mathrm{Pois}(\Lambda) \nonumber \\
N_{it} \sim \mathrm{NB}(\Lambda, \alpha)
\label{eq:N1}
\end{gather}
where $N_{i1}$ is the abundance at site $i=1,\hdots,R$ during year 1
and $\Lambda=\mathbb{E}[N_{i1}]$ is the expected abundance. The negative
binomial model includes the overdispersion parameter $\alpha$.  The
survival and recruitment terms allow abundance to change over time:
\begin{equation}
\left.\begin{aligned}
S_{it}|N_{it-1} &\sim \mathrm{Bin}(N_{it-1}, \omega) \\
G_{it}|N_{it-1} &\sim \mathrm{Pois}(\gamma(N_{it-1})) \\
N_{it} &= S_{it}+G_{it}
\end{aligned}\right\} \quad \text{for} \; t=2,\hdots,T
\label{eq:Nt}
\end{equation}
where $\omega$ is the survival probability and $\gamma$ is the arrival
rate (which can depend on $N_{it-1}$). Dail and Madsen propose three models
for $\gamma$: the constant model, where $\gamma$ does not depend on
$N_{it-1}$, and which simulates a ``propagule rain'' of recruitment; the
autoregressive model , which
simulates geometric or density independent growth; and the ``no-trend''
model, which keeps abundance fixed over time.

>>>>

\subsection{The Data}

The DM model requires count data collected at $R$ sites, each of which
is surveyed on $T$ time periods, such as years. A site is ideally, but
not necessarily, a
well-defined region of a study area, such as a wetland or a patch of
early-successional habitat. In some cases, a site may not have a clear
biological significance, for instance in the case of randomly located
survey plots. The timeframe of the study is arbitrary, but the general
situation is one in which the time period is sufficiently long such
that abundance changes at each site. We will call
these time periods primary sampling periods to distinguish them from
secondary sampling periods, which are repeated surveys within a
primary period. In the case where no secondary sampling was conducted,
let $n_{it}: i=1,\hdots,R; t=1,\hdots,T$ denote the count data at site
$i$ and primary period $t$. If $J$ secondary sampling periods were used, $n_{ijt}:
j=1,\hdots,J$ is the count at site $i$ during secondary sampling
occasion $j$ within year $t$. These are the observed data, which
typically will less than the actual quantitiy of interest, abundance,
denoted $N_{it}$. In cases where detection probability is perfect,
abundance is observed directly such that $\{N_{it}\}$ are the data.

\subsection{The Model}

As originally formulated, the DM model includes submodels for four
conditionally related processes: the initial abundance state, apparent
survival (the number of individuals that survive and do not
emmigrate), recruitment (the number of new individuals entering
the population), and detection probability \citet{dail_madsen:2011}.
Taken together, the first 3 processes describe the state process---the
variation in abundance in space and time. The forth item in the list
is the observation process, which describes the relationship between
the true state, abundance, and the observed count data.

\subsubsection{Initial abundance}

In the original formulation,
initial abundance---abundance at site $i$ during primary period 1, was
modeled using either the Poisson or negative binomial distributions:
\begin{gather}
N_{i1} \sim \mathrm{Pois}(\Lambda_i) \nonumber \\
N_{i1} \sim \mathrm{NB}(\Lambda_i, \alpha)
\label{eq:N1}
\end{gather}
where
%$N_{i1}$ is the abundance at site $i=1,\hdots,R$ during year 1 and
$\Lambda_i$ is the expected abundance at site $i$ during year 1. The
Poisson distribution assumes that the mean of $N_{i1}$ is equal to its
variance---an assumption relaxed when using the negative binomial
distribution, which includes the overdispersion parameter $\alpha$.

In either case, the model for initial abundance has two distinguising
features. First, it allows one to model spatial variation in
abundance. For instance, one might consider the influence of some
environmental covariate on abundance using a simple generalized linear
model such as $\log(\Lambda_i) = \beta^{\Lambda}_0 + \beta^{\Lambda}_1
x_{i}$ which is a log-linear function with an intercept and slope
parameter for the effect of $x_i$ on the logarithm of $\Lambda$.

The second thing to note is that the spatial replicates resolve the
problem of parameter non-identifiability that are common to standard
state-space models.

\subsubsection{Abundance after $t=1$}

The survival and recruitment terms allow abundance to change over time:
\begin{equation}
\left.\begin{aligned}
S_{it}|N_{it-1} &\sim \mathrm{Bin}(N_{it-1}, \omega) \\
G_{it}|N_{it-1} &\sim \mathrm{Pois}(\gamma(N_{it-1})) \\
N_{it} &= S_{it}+G_{it}
\end{aligned}\right\} \quad \text{for} \; t=2,\hdots,T
\label{eq:Nt}
\end{equation}
where $\omega$ is the survival probability and $\gamma$ is the arrival
rate (which can depend on $N_{it-1}$). Dail and Madsen propose three models
for $\gamma$: the constant model, where $\gamma$ does not depend on
$N_{it-1}$, and which simulates a ``propagule rain'' of recruitment; the
autoregressive model , which
simulates geometric or density independent growth; and the ``no-trend''
model, which keeps abundance fixed over time.
